Linear vertical shear instability in astrophysical disks: a semi-numerical study

### 2014-12-20 linear_adia.tex
corrected beta vs nz estimate, updated numerical adiabatic section

\subsubsection{Qualitative expectations in the presence of thermal
  relaxation}
We have shown that when $\gamma>1$ so that there is a positive
vertical entropy gradient and perturbations evolve
adiabatically ($t_c\to\infty$) that the VSI is strongly
surpressed. This is the limit where the term proportional to
$t_c^{-1}$ in the linearized energy equation (Eq. \ref{lin_energy})
becomes negligible compared to the term proportional to
$\gamma/\Gamma-1$. The latter term is associated with the vertical
entropy gradient, or buoyancy frequency $N_z$.

Reversing the argument, we expect to recover the VSI when the bouyancy
frequency becomes negligible compared to the cooling rate, or 
\begin{align}
  t_c \ll \frac{1}{N_z}. 
\end{align} 
Parametrizing $t_c = \beta\Omega_k^{-1}$ and inserting the expression
for $N_z$ appropriate for vertically isothermal disks, we expect
\begin{align}
  \beta \ll
  \frac{1}{H_\mathrm{iso}\sqrt{\frac{\gamma-1}{\gamma}}\left|\frac{\p\ln\rho}{\p
          z}\right|} 
\end{align}
is required to enable the VSI. This inequality is satisfied at the
midplane and as $|z|\to\infty$. It can be satisifed at all heights if
it is satisfied at the height of $\mathrm{max}|\p\ln\rho/\p z|$. 
Using the expression for the density field for vertically isothermal
disks (Eq. \ref{vertiso_eqm}), we expect 
\begin{align}
  \beta \ll \epsilon\frac{3^{3/2}}{2}\sqrt{\frac{\gamma}{\gamma-1}},
\end{align}
i.e. $\beta\ll O(\epsilon)$, is a sufficient condition for
bouyancy to be negligible.

>>>>

\subsubsection{Qualitative expectations in the presence of thermal
  relaxation}
We have shown that when $\gamma>1$ so that there is a positive
vertical entropy gradient and perturbations evolve
adiabatically ($t_c\to\infty$) that the VSI is strongly
surpressed. This is the limit where the term proportional to
$t_c^{-1}$ in the linearized energy equation (Eq. \ref{lin_energy})
becomes negligible compared to the term proportional to
$\gamma/\Gamma-1$. The latter term is associated with the vertical
entropy gradient, or buoyancy frequency squared, $N_z^2$.

Reversing the argument, we expect to recover the VSI when the bouyancy
frequency becomes negligible compared to the cooling rate, or 
\begin{align}
  t_c \ll \frac{\Omega}{\gamma N_z^2}. 
\end{align} 
Parametrizing $t_c = \beta\Omega_k^{-1}$ and inserting the expression
for $N_z^2$ appropriate for vertically isothermal disks, we expect
\begin{align}
  \beta \ll
  \frac{1}{H^2_\mathrm{iso}\left(\gamma-1\right)\left(\frac{\p\ln\rho}{\p
          z}\right)^2} 
\end{align}
is required to enable the VSI, where we used $\Omega\sim\Omega_k$.  
This inequality is satisfied at the
midplane and as $|z|\to\infty$. It can be satisifed at all heights if
it is satisfied at the height of $\mathrm{max}|\p\ln\rho/\p z|$. 
Using the full expression for the density field for vertically isothermal
disks (Eq. \ref{vertiso_eqm}), we expect 
\begin{align}
  \beta \ll \frac{27}{4(\gamma-1)}\epsilon^2,
\end{align}
i.e. $\beta\ll O(\epsilon^2)$, is a sufficient condition for
bouyancy to be negligible.

### 2014-12-20 numerical.tex
corrected beta vs nz estimate, updated numerical adiabatic section

Unless otherwise stated, our fiducial disk model is a
nearly-vertically isothermal disk with $\Gamma=1.011$, $p=-1.5$,
$q=-1$, $\epsilon=0.1$, and the perturbation wavenumber $k_x = 200\pi$
(or $\hat{k}=20\pi$).

@@

Fig. \ref{lowfreq_eigen_adia} show the eigenvalues for this
problem. The growth rates are about an order of magnitude smaller than
the previous case. The fundamental VSI mode is that with
$\mathrm{max}|\nu|$. Its expected and numerically-calculated growth
rates compares well, with 
\begin{align*}
  \nu = 0.03751 \epsilon \Omega_k &\quad \text{(from
    Eq. \ref{gam2_growth_rate})},\\
  \nu = 0.03672 \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
The fundamental mode is plotted in Fig. \ref{lowfreq_eigenfunc_adia},
and clearly show that perturbations are rapidly stabilized away from
the midplane. For the adopted disk and perturbation parameters 
($|q|=1$ and $\hat{k} = 20\pi$), Eq. \ref{gam2_alpha} gives $\real{\alpha}\simeq - 43.8$,
corresponding to a characteristic decay lengthscale of $\simeq
0.15\epsilon r$, which is consistent with that observed in 
Fig. \ref{lowfreq_eigenfunc_adia}.

>>>>

Unless otherwise stated, our fiducial disk model is a
nearly-vertically isothermal disk with $\Gamma=1.011$, $p=-1.5$,
$q=-1$, $\epsilon=0.1$, and the perturbation wavenumber $k_x = 200\pi/r$
(or $\hat{k}=20\pi$).

@@

Fig. \ref{lowfreq_eigen_adia} show the eigenvalues for this
problem. The growth rates are about an order of magnitude smaller than
the previous case. The fundamental VSI mode is that with
$\mathrm{max}|\nu|$. Its expected and numerically-calculated growth
rates compares well, with 
\begin{align*}
  \nu = 0.03751 \epsilon \Omega_k &\quad \text{(from
    Eq. \ref{gam2_growth_rate})},\\
  \nu = 0.03672 \epsilon \Omega_k &\quad \text{(numerical)}.
\end{align*}
The fundamental mode is plotted in Fig. \ref{lowfreq_eigenfunc_adia},
and clearly show that perturbations are rapidly stabilized away from
the midplane. For the adopted disk and perturbation parameters 
($|q|=1$ and $\hat{k} = 20\pi$), Eq. \ref{gam2_alpha} gives $\real{\alpha}\simeq - 43.8$,
corresponding to a characteristic decay length scale of $\simeq
0.15\epsilon r$, which is consistent with that observed in 
Fig. \ref{lowfreq_eigenfunc_adia}.

Note that the decay lengthscale is rougly the height within which
vertical shear is larger than the buoyancy frequency squared. That is,
\begin{align*}
\left|r\frac{d\Omega^2}{dz}\right|\gtrsim N_z^2
\end{align*}
for
\begin{align*}
  \left|\frac{z}{\epsilon r}\right| \lesssim \frac{\epsilon|q|\gamma}{\gamma-1},
\end{align*}
in a thin, vertically isothermal disk. This evaluates to $z\sim
0.2\epsilon r$ for the current disk model, consistent with
Fig. \ref{lowfreq_eigenfunc_adia}.

Furthermore, the growth rate to should be no larger than the
maximum vertical shear available within this decay lengthscale. Using 
Eq. \ref{max_growth} as a rough guide, 
\begin{align}
  \frac{\nu}{\epsilon\Omega_k}<\frac{\epsilon |q|^2}\gamma{2(\gamma-1)},
\end{align}
or $\nu< 0.1\epsilon\Omega_k$ is expected in our disk model,
consistent with numerical results.
